---
title: "Lecture 6.2 TEMPLATE"
subtitle: "Annotations"
author: "Your Name"
date: 2025-02-10
toc: true
---

::: {.callout-note}
This template follows [lecture 6.2 slides](https://eds-240-data-viz.github.io/course-materials/lecture-slides/lecture6.2-annotations-slides.html#/title-slide){target="_blank"}. Please be sure to cross-reference the slides, which contain important information and additional context!
:::

## Lobster plot

### Setup

```{r}
#| eval: true
#| echo: true
#| warning: false
#| message: false
#.........................load libraries.........................
library(tidyverse)

#..........................read in data..........................

# read in Google Sheet ----
lobs <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1DkDVcl_9rlaqznHfa_v1V1jtZqcuL75Q6wvAHpnCHuk/edit#gid=2143433533") |>
  mutate(temp = as.factor(temp))
```

### Create base plot

```{r}
#| eval: true
#| echo: true
#| warning: false
#..........................create theme..........................
lob_theme <- function(){
  theme_light() +
    theme(
      axis.title.x = ggtext::element_markdown(size = 16,
                                              lineheight = 1.5,
                                              margin = margin(t = 10, r = 0, b = 0, l = 0)),
      axis.title.y = ggtext::element_markdown(size = 16,
                                              lineheight = 1.5,
                                              margin = margin(t = 0, r = 10, b = 0, l = 0)),
      axis.text = element_text(color = "black", 
                               size = 14),
      axis.ticks = element_line(color = "black"),
      panel.border = element_rect(colour = "black", 
                                  linewidth = 0.7),
      panel.grid = element_blank(),
      legend.title = element_text(size = 15),
      legend.text = element_text(size = 13),
      legend.position = c(0.95, 0.95),
      legend.justification = c(0.95, 0.95),
      legend.box.background = element_rect(linewidth = 0.9)
    )
}

#..........................create scales.........................
lob_palette <- c("11" = "#7B8698",
                 "16" = "#BAD7E5",
                 "21" = "#DC7E7C",
                 "26" = "#7D3E40")

lob_shapes <-  c("11" = 15,
                 "16" = 16,
                 "21" = 17,
                 "26" = 18)

lob_sizes <- c("11" = 6,
               "16" = 6,
               "21" = 6,
               "26" = 7)

#........................create plot text........................
x_axis_lab <- glue::glue("Resting Metabolic Rate<br>
                         (mg O<sub>2</sub> kg<sup>-1</sup> min<sup>-1</sup>)")

y_axis_lab <- glue::glue("Maximum Consumption Rate<br>
                         (prey consumed predator<sup>-1</sup> 24hr<sup>-1</sup>)")

#............................plot data...........................
lob_plot <- ggplot(lobs, aes(x = SMR, y = avg_eaten,
                             color = temp, shape = temp, size = temp)) +
  geom_point() +
  scale_color_manual(values = lob_palette, name = "Temperature (ºC)") +
  scale_shape_manual(values = lob_shapes, name = "Temperature (ºC)") +
  scale_size_manual(values = lob_sizes, name = "Temperature (ºC)") +
  scale_x_continuous(breaks = seq(0, 1.5, by = 0.2)) +
  scale_y_continuous(breaks = seq(0, 35, by = 5)) +
  labs(x = x_axis_lab,
       y = y_axis_lab) +
  lob_theme()

lob_plot
```

### Adding annotations using `geom_text()` & `geom_rect()`

- bounding box + text

```{r}
#| eval: true
#| echo: true
#| warning: false
# create bounding box, text ----
lob_plot + 
```

### Adding annotations using `annotate()`

- bounding box + text + arrow

```{r}
#| eval: true
#| echo: true
#| warning: false
# create bounding box, text, arrow ----
lob_plot +
```

### Annotate each point

```{r}
#| eval: true
#| echo: true
# using geom_text() ----
lob_plot +

# using geom_label() ----
lob_plot +
  geom_label(aes(label = lobster_id),
             size = 6,
             show.legend = FALSE)

# using ggrepel (text) ----
lob_plot +
  
# using ggrepel (labels) ----
lob_plot +
  ggrepel::geom_label_repel(aes(label = lobster_id),
                           size = 4,
                           color = "gray10",
                           nudge_x = 0.1, nudge_y = 0.3,
                           arrow = arrow(length = unit(0.25, "cm")))
```

### Use `annotate()` to label a few points

- ID + arrow

```{r}
#| eval: true
#| echo: true
# label 2 points with lobster ID & arrow ----
lob_plot +
```

## Mono Lake plot

### Setup

```{r}
#| eval: true
#| echo: true
#| message: false
#..........................read in data..........................

# read in Google Sheet ----
mono <- googlesheets4::read_sheet("https://docs.google.com/spreadsheets/d/1o0-89RFp2rI2y8hMQWy-kquf_VIzidmhmVDXQ02JjCA/edit#gid=164128885")
```

### Create base plot 

```{r}
#| eval: true
#| echo: true
# create base plot ----
ggplot(data = mono, aes(x = year, y = lake_level)) +
  geom_line() +
  labs(x = "\nYear",
       y = "Lake surface level\n(feet above sea level)\n",
       title = "Mono Lake levels (1850 - 2017)\n",
       caption = "Data: Mono Basin Clearinghouse") +
  scale_x_continuous(limits = c(1850, 2020),
                     expand = c(0,0),
                     breaks = seq(1850, 2010, by = 20)) +
  scale_y_continuous(limits = c(6350, 6440),
                     breaks = c(6370, 6400, 6430),
                     expand = c(0,0),
                     labels = scales::label_comma()) +
  theme_light() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(face = "italic")
    )
```

### Add annotations

- order of annotation layers matters!

```{r}
#| eval: true
#| echo: true
ggplot(data = mono, aes(x = year, y = lake_level)) +
  geom_line() +
  labs(x = "\nYear",
       y = "Lake surface level\n(feet above sea level)\n",
       title = "Mono Lake levels (1850 - 2017)\n",
       caption = "Data: Mono Basin Clearinghouse") +
  scale_x_continuous(limits = c(1850, 2020),
                     expand = c(0,0),
                     breaks = seq(1850, 2010, by = 20)) +
  scale_y_continuous(limits = c(6350, 6440),
                     breaks = c(6370, 6400, 6430),
                     expand = c(0,0),
                     labels = scales::label_comma()) +
  theme_light() +
  theme(
    plot.title.position = "plot",
    plot.title = element_text(size = 16),
    axis.title = element_text(size = 12),
    axis.text = element_text(size = 10),
    plot.caption = element_text(face = "italic")
    )
```
